### 2020-07-14 reproduction/Tabs/tab2.tex
added reproduction results; comments in reproduction.rmd

% latex table generated in R 3.5.1 by xtable 1.8-4 package
% Mon Jul 13 19:13:01 2020
\begin{tabular}{cccr@{--}lr@{--}l}
  \hline
  & \multicolumn{2}{c}{Estimate (\%)} & \multicolumn{4}{c}{95\% credible interval}\\ Rule & original & reproduction & \multicolumn{2}{c}{original} & \multicolumn{2}{c}{reproduction}\\ \hline
100\% & 34 & 34 & 27 & 39 & 27 & 39 \\ 
  90\% & 36 & 36 & 29 & 42 & 29 & 42 \\ 
  5'80\%-3'100bp & 35 & 35 & 28 & 41 & 28 & 41 \\ 
  80\% & 40 & 40 & 33 & 46 & 33 & 46 \\ 
  70\% & 42 & 42 & 35 & 49 & 35 & 49 \\ 
  60\% & 42 & 42 & 33 & 50 & 33 & \textbf{\color{red} 49} \\ 
   \hline
\end{tabular}

### 2020-07-14 reproduction/reproduction.Rmd
added reproduction results; comments in reproduction.rmd

---
title:  "Reproduction of Lamichhane et al. (2003)"
author: "Karl Broman"
date:   "`r Sys.Date()`"
output:
    html_document:
        code_folding: hide
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r set_knitr_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

This is a reproduction of the results in Lamichhane et al.
(2003) A post-genomic method for predicting essential genes at
subsaturation levels of mutagenesis: application to _Mycobacterium
tuberculosis_. Proc Natl Acad Sci USA 100:7213-7218
[doi:10.1073/pnas.1231432100](https://doi.org/10.1073/pnas.1231432100)
The full repository for the reproduction work is
<https://github.com/kbroman/Paper_ReScience2020>.

The original project directory, with data and code, is on
[github](https://github.com/kbroman/Project_Lamichhane2003). A subset
of the materials is in [`../original/`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original), and I will be
drawing from that. The code to create Figure 1b in the paper
was not in that project directory but rather with materials for a talk
I gave in 2002-2003. The materials for that talk is on
[github](https://github.com/kbroman/Talk_Mtb). A subset is in
[`../talk/`](https://github.com/kbroman/Paper_ReScience2020/blob/master/talk).

I should say right at the start that _I am not able to reproduce
Figure 3_ in the paper. I have the code to make the Figure based on
simulation results, but I can't find the code that was used to carry
out the simulation.

I will first load the required libraries for this analysis.

```{r load_libraries}
library(negenes)
library(R.utils)  # for gunzip()
library(devtools) # for session_info()
library(gt)       # for tables
library(xtable)   # for latex tables
library(readxl)   # for reading excel file
library(broman)   # for grayplot()
```

## Move stuff over

The first task is to copy over the primary data and scripts that we
need to reproduce the results. I will first create a set of
subdirectories to contain them.

```{r create_subdir}
dir <- c("Rawdata", "Rawdata/TIGR", "Data", "R", "Figs", "Tabs")

for(d in dir) {
    if(!dir.exists(d)) dir.create(d)
}
```

Next, I will copy over the primary data files.

```{r copy_rawdata}
raw_data_files <- c("Rawdata/TIGR/GMT.1con.gz",
                    "Rawdata/TBCDC1551_rev.csv",
                    "Rawdata/MTCoords_rev.csv",
                    "Rawdata/MT-RvConversion_rev.csv",
                    "Rawdata/MtbGeneClassification.csv",
                    "Sept02/Rawdata/GenomicData-final4_rev.csv",
                    "Sept02/Rawdata/Phase2-FinalData_rev.csv")

for(file in raw_data_files) {
    orig_file <- file.path("..", "original", file)
    new_file <- sub("Sept02/", "", file, fixed=TRUE)

    if(!file.exists(new_file)) {
        cat(orig_file, " -> ", new_file, "\n")
        file.copy(orig_file, new_file)
    }
}
```

I need to use `gunzip()` to unzip the TIGR genomic data file.

```{r unzip_tigr_file}
gzipped_file <- "Rawdata/TIGR/GMT.1con.gz"
gunzipped_file <- sub(".gz", "", gzipped_file, fixed=TRUE)
if(!file.exists(gunzipped_file)) {
    gunzip(gzipped_file)
}
```

I will now copy over the perl script that pulls out the transposon
insertion sites from the _M. tuberculosis_ genome.

```{r copy_perl}
file <- "findTA.pl"
if(!file.exists(file)) {
    file.copy(file.path("..", "original", file), file)
}
```

Now I will copy over the R scripts that do the MCMC analysis.
The ones in `Sept02/R` create results that appear in Table 2. The ones
in `Nov02/R` create all of the rest of the results.

```{r copy_r_scripts}
files <- c("prepareData.R", "analysis.R")

# 2002-09 versions
for(file in files) {
    orig_file <- file.path("..", "original", "Sept02", "R", file)
    new_file <- file.path("R", paste0("sept02_", file))

    if(!file.exists(new_file)) {
        file.copy(orig_file, new_file)
    }
}

# 2002-11 versions
for(file in files) {
    orig_file <- file.path("..", "original", "Nov02", "R", file)
    new_file <- file.path("R", paste0("nov02_", file))

    if(!file.exists(new_file)) {
        file.copy(orig_file, new_file)
    }
}
```

Finally, I will copy over the R script that create the figures.

```{r copy_fig_script}
file <- "figs4paper.R"
orig_file <- file.path("..", "original", "Nov02", "R", file)
new_file <- file.path("R", file)
if(!file.exists(new_file)) {
    file.copy(orig_file, new_file)
}
```

Oh, yeah. There is also the script that makes Figure 1b,
which was not in the original project directory but rather with
materials for a talk I gave at the time.

```{r circle_fig_script}
file <- "circlefig.R"
orig_file <- file.path("..", "talk", "R", file)
new_file <- file.path("R", file)
if(!file.exists(new_file)) {
    file.copy(orig_file, new_file)
}
```

## Run perl script

I will first run the Perl script `findTA.pl` which uses data from
`Rawdata/` and creates a set of files in `Data/`.

```{r run_perl_script}
system("perl findTA.pl")
```

*LW: It might be helpful to explain here (more specifically) what 
the Perl script does? Also, I'm not sure if you want to include the 
commandline printouts. They are always consistent for me so maybe 
this is not very important.*

## Project structure

The reproduction of this work included a number of challenges: lack of
documentation (couldn't you give me even one README file, 2002 Karl?),
multiple directories of scripts with portions of results coming from
different ones, not saving random number generator seeds,
not loading the required library in the scripts,
not saving intermediate results to files (but just leaving them in the
R environment, in `.RData`), and not explaining which R objects were
needed in a script.

To reproduce the work, I am going to run a particular script, save the
key results to a file, and then clean the R environment for the next
step. Some of the calculations take a bit of time, and those I will
cache using a crude caching system that will mostly be hidden.

Because of the lack of documentation, the first challenge was to
figure out which results I was trying to reproduce. To do that, I
re-read the paper. The key results are in Figures 1, 2, and 3, and
Tables 2, 3, and 4. Table 3 also has an extended version, as
Supplementary Table 6. There are a number of
other summary statistics about the data, for example in Table 1 and in
the right-hand column of the second page of the paper. But I am going
to focus on the key analysis results, Figures 1-3 and Tables 2-4 and 6.

Let me first explain some key elements of the project directory in
[`../original/`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/). The `R` subdirectory contains my
initial analysis of the data (performed in like July, 2002), which did
not end up in the paper. [`Sept02/R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Sept02/R) contains a
revised analysis based on a slightly larger set of data and evaluating multiple
versions of a rule for part of the data handling (see Table 2
of the published paper).
[`Nov02/R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R) contains an analysis with our final
version of the data handling rule and is the basis for one row of
Table 2 and all of the rest of the analyses.

*LW: I think this chunk should be moved to the top before all the file
renaming happened. The explanation for Nov02 being "the basis for one row 
of Table 2 and all of the rest of the analyses” is a bit confusing to me. 
I think instead of making vague gestures like so, it would be more helpful
to clarify “Nov02 adds an extra parameter” instead of “adds an extra row” 
or whatever you meant to say here.*

## Run MCMC

The analysis proceeded in two steps: run a `prepareData.R` script to
create the key data objects, and then run `analysis.R`
which performs the Markov chain Monte Carlo (MCMC) analysis and summarizes
the results. We first need to do the `Sept02` versions and then the
`Nov02` versions.

So we will start with
[`Sept02/R/prepareData.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Sept02/R/prepareData.R), which
we have saved as [`R/sept02_prepareData.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/reproduction/R/sept02_prepareData.R). It
prints a bunch of crap, so we will hide the output.

There is one small wrinkle here. The `ngeneperfam` object has
class `"table"` which leads to an error in the next step, in the
[`analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Sept02/R/analysis.R) script. This is due to
some change in [R](https://www.r-project.org) since 2002. So we need
to convert the `ngeneperfam` object to a simple numeric vector.

```{r run_sept02_prepareData, results="hide"}
rm(list=ls())
setwd("R")
source("sept02_prepareData.R")

# need to turn this "table" into numeric
ngeneperfam <- setNames(as.numeric(ngeneperfam), names(ngeneperfam))

# save the data objects to a file
save(data100nostop, data100stop, data90, data70, data60, data80,
     ngeneperfam, numTAs, geneclasses,
     file="sept02_data.RData")

setwd("..")
```

*LW: I understand resetting the wd in each code chunk is necessary for the R scripts sourced
to run correctly, but I still think modifying those R scripts to all use the same reference
wd ('/reproduction in this case') would make it a lot more clear and consistent.*

I will now run the MCMC. (I actually cache these results and only run
them if needed.) Again, we copied over the
[`Sept02/R/analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Sept02/R/analysis.R) script as
[`R/sept02_analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/reproduction/R/sept02_analysis.R).

The only slight change here is to rename one of the
objects to avoid conflict with the results obtained using the
[`Nov02`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R) script, below.

```{r run_sept02_mcmc, eval=FALSE}
rm(list=ls())
setwd("R")

# load the data created above
load("sept02_data.RData")

set.seed(74250112)
source("sept02_analysis.R")

# A different finalres object is created by the Nov02 script
# so change the name of this thing to res80
# (but note that res80 is in different format from the others)
res80 <- finalres
save(res100nostop, res90, res80, res70, res60,
     file="sept02_results.RData")

setwd("..")
```

*LW: since MCMC here takes a good while to run, I think it might be helpful to include
a note reminding the people using the code or include a sample run time so people have 
a rough idea of how long it might take.*

```{r really_run_sept02_mcmc, echo=FALSE, results="hide"}
if(file.exists("R/sept02_results.RData")) {
    load("R/sept02_results.RData")
} else {
<<run_sept02_mcmc>>
}

# LW: "<<>>" syntax doesn't work in rstudio. Might be markdown specific?
```

We now repeat the same things with the [`Nov02`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R)
scripts. First,
[`Nov02/R/prepareData.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R/prepareData.R), which we
have saved as [`R/nov02_prepareData.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/reproduction/R/nov02_prepareData.R). Again,
it prints a bunch of crap, so we will hide the output.

Again, we need to convert the `ngeneperfam` object to be a
numeric vector.

```{r run_nov02_prepareData, results="hide"}
rm(list=ls())
setwd("R")
source("nov02_prepareData.R")

# need to turn this "table" into numeric
ngeneperfam <- setNames(as.numeric(ngeneperfam), names(ngeneperfam))

# save the data objects to a file
save(mydata,
     ngeneperfam, numTAs, geneclasses,
     taloc, classes,
     file="nov02_data.RData")

setwd("..")
```

And again we run the MCMC. This one is faster, because we are doing
just one version of the analysis rather than 6. The
[`R/nov02_analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/reproduction/R/nov02_analysis.R) script was copied over from
[`Nov02/R/analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R/analysis.R).

The only modification here is to change the name of one of the results
objects, from `wh` (which is not very meaningful) to `genes`.

```{r run_nov02_mcmc, eval=FALSE}
rm(list=ls())
setwd("R")

# load the data created above
load("nov02_data.RData")

set.seed(27150742)
source("nov02_analysis.R")

genes <- wh # rename "wh" (as more meaningful)
save(finalres, fams, genes, famprob,
     file="nov02_results.RData")

setwd("..")
```

```{r really_run_nov02_mcmc, echo=FALSE, results="hide"}
if(file.exists("R/nov02_results.RData")) {
    load("R/nov02_results.RData")
} else {
<<run_nov02_mcmc>>
}
```

## Tables

Let us now reconstruct the tables 2-4 and 6.

### Table 2

First, let us look at Table 2, which shows the estimated percent
essential genes (with an interval estimate) using various rules for
defining the part of the gene where viable transposon insertion would
indicate that the gene is not essential.

```{r table2}
rm(list=ls())
load("R/sept02_results.RData")
load("R/nov02_results.RData")

tab2 <- matrix(nrow=6, ncol=3)
dimnames(tab2) <- list(c("100%", "90%", "5'80%-3'100bp", "80%", "70%", "60%"),
                       c("estimate (%)", "lo 95% CI", "hi 95% CI"))

for(i in 1:6) {
    # grab the appropriate result object
    obj_nam <- c("res100nostop", "res90", "finalres",
                 "res80", "res70", "res60")[i]
    obj <- get(obj_nam)[[1]] # first component is number of essential genes

    # pull out the summaries
    # divide by the number of genes (4204) and multiple by 100 (to get percent)
    tab2[i,] <- c(mean(obj), quantile(obj, c(0.025, 0.975))) / 4204 * 100
}

# add rule as a first column
tab2 <- cbind(data.frame(rule=rownames(tab2)), tab2)

# use gt package to make a table
gt(tab2) %>%
    fmt_number(columns=2:4, decimals=0)
```

Comparing the reconstructed table to the [published
version](https://bit.ly/lamichhane2003_table2),
we see just one small difference:
the upper confidence limit in the bottom row is 49% here,
whereas it was 50% in the published paper. This difference can be
ascribed to MCMC sampling variation.

The following creates a LaTeX table comparing these results, for the
final paper.

```{r create_table2}
tab2_new <-
    cbind(tab2[,1,drop=FALSE], estimate_orig=c(34,36,35,40,42,42),
          round(tab2[,2,drop=FALSE]),
          ci_lo_orig=c(27,29,28,33,35,33),
          ci_hi_orig=c(39,42,41,46,49,50),
          round(tab2[,3:4]), stringsAsFactors=FALSE)

# highlight differences in bold [there is just one]
for(i in 1:nrow(tab2_new)) {
  if(tab2_new[i,2] != tab2_new[i,3])
      tab2_new[i,3] <- paste0("\\textbf{\\color{red} ", tab2_new[i,3], "}")
  if(tab2_new[i,4] != tab2_new[i,6])
      tab2_new[i,6] <- paste0("\\textbf{\\color{red} ", tab2_new[i,6], "}")
  if(tab2_new[i,5] != tab2_new[i,7])
      tab2_new[i,7] <- paste0("\\textbf{\\color{red} ", tab2_new[i,7], "}")
}

xtab2 <- xtable(tab2_new, digits=0,
               align=c(rep("c", 4), rep(c("r@{--}", "l"), 2)))

addtorow <- list()
addtorow$pos <- list(0,0)
addtorow$command <- c("& \\multicolumn{2}{c}{Estimate (\\%)} & \\multicolumn{4}{c}{95\\% credible interval}\\\\",
                      "Rule & original & reproduction & \\multicolumn{2}{c}{original} & \\multicolumn{2}{c}{reproduction}\\\\")

print(xtab2, file="Tabs/tab2.tex",
      include.rownames=FALSE,
      include.colnames=FALSE,
      sanitize.text.function=function(x) gsub("%", "\\%", x, fixed=TRUE),
      add.to.row=addtorow,
      floating=FALSE)
```

### Tables 3 and 6

Let us now turn to Tables 3 and 6, on the genes with probability > 75%
of being essential.

In order to reproduce these tables exactly, we need to make one small
change to the code in the
[`Nov02/R/analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R/analysis.R).
While the original script pulled out genes with probability &ge;
0.749, we need to change that cutoff to 0.745. This difference is
within rounding error and can be ascribed to MCMC sampling error.

We can make the change by reading in the script, pulling out the
appropriate lines, substituting the one number, and then re-running those
lines.

```{r construct_table3_and_6}
rm(list=ls())
load("R/nov02_data.RData")    # numTAs
load("R/nov02_results.RData") # finalres

# read script; subset to lines 50-53; change 0.749 to 0.745
script <- readLines("R/nov02_analysis.R")
lines <- sub("0.749", "0.745", script[50:53], fixed=TRUE)

# run those lines
eval(parse(text=lines))

# add gene number as a column; change row order; change name of object
wh <- cbind(data.frame(MTnum=rownames(wh)), wh)
wh <- wh[rev(order(wh[,3], -as.numeric(rownames(wh)))), ]
wh[14:15,] <- wh[15:14,]
wh[19:20,] <- wh[20:19,]
# I am not quite seeing the point of the two lines above. 
# The reverse order does not apply to all rows with the same probability.
# It seems very particular to this dataset. Is it just to make the table
# here look the exact same as the one in the original publication?

genes <- wh
genes[,3] <- genes[,3] * 100 # convert to percent
```

*LW: I got confused when I was running through it because the first table generated here is
in fact supplement table 6 and not table 3. I think the legend of table 2 in the 2020 paper
if added here would make a lot more sense about the distinction between table 3 and table s6.*

Here is the reconstruction of Table 3. The results match the [published
version](https://bit.ly/lamichhane2003_table3).
(If we had left the cutoff at 0.749, we would have missed the
last two genes on the list, for which the estimated probability of
being essential is now 74.8%. This difference is again likely due to
MCMC sampling error.)

```{r table3}
# genes to exclude from Table 3 (just listed in table 6)
tab3_exclude <- c("418",  "1218", "3003", "1701",
                  "2448", "3002", "1702")

genes[!(genes$MTnum %in% tab3_exclude), c("MTnum", "Prob.essential")] %>%
    gt() %>%
    fmt_number(columns="Prob.essential", decimals=0)
```

Seven genes were excluded from Table 3 (because there was other
evidence to indicate that they were non-essential), but the full list
was shown in Supplemental Table 6. Here is the reconstruction of the
full table, which matches the published version.

```{r table6}
genes[, c("MTnum", "Prob.essential")] %>%
    gt() %>%
    fmt_number(columns="Prob.essential", decimals=0)
```

The following creates a LaTeX table comparing these results, for the
final paper.

```{r create_table6}
# download original table 6
url <- "https://www.pnas.org/highwire/filestream/586425/field_highwire_adjunct_files/1/1432Table6.xls"
local_file <- "Tabs/1432Table6.xls"
if(!file.exists(local_file)) {
    download.file(url, local_file)
}

# LW: link download didn't work for me. I downloaded the file into the directory myself. Would it be better if it is just included in a directory (on GitHub) to start out with?

# read original table 6
tab6_orig <- readxl::read_excel(local_file, sheet=1, skip=2)
tab6_orig <- as.data.frame(tab6_orig)

# add gene MT number
genes$MTnum <- sprintf("%04d", as.numeric(as.character(genes$MTnum)))

stopifnot(all(tab6_orig[,1] == genes$MTnum))

tab6_new <- cbind(tab6_orig, round(genes$Prob.essential))

# highlight differences in bold [there aren't any]
for(i in 1:nrow(tab6_new)) {
    if(tab6_new[i,4] != tab6_new[i,5])
        tab6_new[i,5] <- paste0("\\textbf{\\color{red} ", tab6_new[i,5], "}")
}

# remove the references at the end of the gene descriptions
tab6_new[,3] <- sub("\\([123]\\)$", "", tab6_new[,3])

xtab6 <- xtable(tab6_new, digits=0, align=rep("c", 6))

addtorow <- list()
addtorow$pos <- list(0,0)
addtorow$command <- c("& & & \\multicolumn{2}{c}{Probability (\\%)}\\\\",
                      "MT \\# & Rv \\# & Gene description & original & reproduction\\\\")

print(xtab6, file="Tabs/tab6.tex",
      include.rownames=FALSE,
      include.colnames=FALSE,
      sanitize.text.function=function(x) gsub("%", "\\%", x, fixed=TRUE),
      add.to.row=addtorow,
      floating=FALSE)
```

### Table 4

Table 4 contains a list of gene families that appear to be enriched
with essential genes, or to have a deficit of essential genes.
The contents of the table are created in
[`Nov02/R/analysis.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R/analysis.R), but one small
change needs to be made, replacing the 75% cutoff with 74.5%, as
otherwise one of the gene families gets left off. (The new estimate of
the enrichment probability for one family has dipped slightly below 75%.)

```{r table4}
# LW: did you miss the "clear environment" line here?
load("R/nov02_data.RData")
load("R/nov02_results.RData")

# read script
script <- readLines("R/nov02_analysis.R")

# change cutoff from 75 to 74.5
lines <- sub(">= 75", ">= 74.5", script[56:64], fixed=TRUE)

# run the lines
eval(parse(text=lines))

# paste in family labels
fams <- cbind(data.frame(family=classes[fams[,1]]),
              fams[,-1])

# create the table
gt(fams[,c("prob.enriched", "family", "percent.essential", "2.5%", "97.5%")])
```

There are a few small changes from the [published
version](https://bit.ly/lamichhane2003_table4). First, the
enrichment probability has changed &plusmn;1% for three of the
families. Second, the upper confidence limits have changed for two
families: in the first row the limit changed from 76 to 72%, and in
the second-to-last row it changed from 36 to 38%. I think these
differences can be ascribed to MCMC sampling error.

The following creates a LaTeX table comparing these results, for the
final paper.

```{r create_table4}
# original table 4
tab4_orig <- data.frame(prob_enriched=c(97,94,82,80,78,75,75,4,4,4,0),
                        percent_essential=c(54,45,46,40,42,43,42,32,20,27,10),
                        ci_essential_lo=c(32,30,21,28,23,21,20,25,7,17,0),
                        ci_essential_hi=c(76,60,68,52,62,64,65,39,40,36,24),
                        group=c("Aminoacyl tRNA synthases and their modification",
                                "PE family: PGRS subfamily",
                                "Purine ribonucleotide biosynthesis",
                                "Polyketide and nonribosomal peptide synthesis",
                                "Synthesis of fatty and mycolic acids",
                                "Ser/Thr protein kinases and phosphoprotein phosphatases",
                                "Biosynthesis of molybdopterin",
                                "Unknown proteins",
                                "Metabolism of sulphur",
                                "PPE family",
                                "Conserved membrane proteins"),
                        stringsAsFactors=FALSE)

# clean up the new names
fams[,1] <- as.character(fams[,1])
fams[,1] <- sub("^[A-Za-z]+: ", "", fams[,1])
fams[fams[,1]=="Unknowns",1] <- "Unknown proteins"
fams[fams[,1]=="Intermediate Metabolism-of sulphur",1] <- "Metabolism of sulphur"
fams[fams[,1]=="PPE family of proteins",1] <- "PPE family"
fams[fams[,1]=="Biosyn of Molybdopterin",1] <- "Biosynthesis of molybdopterin"
fams[,1] <- sub("non-ribosomal", "nonribosomal", fams[,1], fixed=FALSE)
fams[,1] <- sub("Biosyn ", "Synthesis ", fams[,1], fixed=FALSE)

# reorder as in original table
m <- match(tab4_orig$group, fams[,1])
stopifnot( !any(is.na(m)) )
fams <- fams[m, ]
stopifnot( all( fams[,1] == tab4_orig$group ) )

tab4_new <- cbind(tab4_orig[,5,drop=FALSE],
                  tab4_orig[,1,drop=FALSE], fams[,2,drop=FALSE],
                  tab4_orig[,2,drop=FALSE], fams[,3,drop=FALSE],
                  tab4_orig[,3:4,drop=FALSE], fams[,4:5,drop=FALSE],
                  stringsAsFactors=FALSE)

# highlight differences in bold
for(i in 1:nrow(tab4_new)) {
    if(tab4_new[i,2] != tab4_new[i,3])
        tab4_new[i,3] <- paste0("\\textbf{\\color{red} ", tab4_new[i,3], "}")
    if(tab4_new[i,4] != tab4_new[i,5])
        tab4_new[i,5] <- paste0("\\textbf{\\color{red} ", tab4_new[i,5], "}")
    if(tab4_new[i,6] != tab4_new[i,8])
        tab4_new[i,8] <- paste0("\\textbf{\\color{red} ", tab4_new[i,8], "}")
    if(tab4_new[i,7] != tab4_new[i,9])
        tab4_new[i,9] <- paste0("\\textbf{\\color{red} ", tab4_new[i,9], "}")
}

# reduce group names
tab4_new[1,1] <- paste0(substr(tab4_new[1,1], 1, 24), "...")
tab4_new[4,1] <- paste0(substr(tab4_new[4,1], 1, 27), "...")
tab4_new[6,1] <- paste0(substr(tab4_new[6,1], 1, 27), "...")

# rearrange columns
tab4_new <- tab4_new[,c(1,2,3,4,6,7,5,8,9)]

tab4_new[,6] <- paste0(tab4_new[,6], ")")
tab4_new[,9] <- paste0(tab4_new[,9], ")")

xtab4 <- xtable(tab4_new, digits=0,
               align=c("c", "l", rep("c", 2), rep(c("c", "@{ (}r@{--}", "l"), 2)))

addtorow <- list()
addtorow$pos <- list(0,0)
addtorow$command <- c(paste("& \\multicolumn{2}{c}{Probability enriched (\\%)} &",
                            "\\multicolumn{6}{c}{Est. \\% essential}\\\\"),
                      paste("Functional group & original & reproduction &",
                            "\\multicolumn{3}{c}{original} & \\multicolumn{3}{c}{reproduction}\\\\"))

print(xtab4, file="Tabs/tab4.tex",
      include.rownames=FALSE,
      include.colnames=FALSE,
      sanitize.text.function=function(x) gsub("%", "\\%", x, fixed=TRUE),
      add.to.row=addtorow,
      floating=FALSE)
```

## Figures

We now turn to reconstruction of the Figures.

### Figure 1b

Let us first consider Figure 1b, which shows the transposon insertion
sites on the circular genome of _M. tuberculosis_.

The code was not contained within the project directory, but I found
it in a separate directory of files for a talk I gave on the work in
2002-2003.

```{r figure1b}
rm(list=ls())
setwd("R")

mygeneloc <- read.csv("../Data/mygeneloc.csv")
load("nov02_data.RData")

source("circlefig.R")

setwd("..")
```

The code creates a file `Figs/circlefig.ps`.  To generate the figure within
the current document, we can read in the code,
comment out the `postscript()` and `dev.off()` lines, and then execute
them, as follows.

```{r figure1b_again, fig.width=6, fig.height=6}
mygeneloc <- read.csv("Data/mygeneloc.csv")
load("R/nov02_data.RData")

# read circlefig code
circlefig <- readLines("R/circlefig.R")
# comment out postscript() and dev.off()
circlefig <- sub("postscript(", "#postscript(", circlefig, fixed=TRUE)
circlefig <- sub("dev.off(", "#dev.off(", circlefig, fixed=TRUE)

# run the code here
eval(parse(text=circlefig))
```

The reconstructed figure appears to be the same as the [published
figure](https://bit.ly/lamichhane2003_fig1).

### Figures 1a and 2

The code to create Figure 1a (with the transposon insertion locations)
and Figure 2 (probability essential vs number of insertion sites, for
each gene) were together in a single file,
[`Nov02/R/figs4paper.R`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R/figs4paper.R).

It is a simple matter to run the script, which generates postscript
files `Figs/fig1.ps` and `Figs/fig2.ps`. Points in Figure 2 with
probability 0 are jittered vertically, so we set the random number
seed here.

```{r figure1a_and_figure2}
rm(list=ls())
load("R/sept02_data.RData")
load("R/nov02_data.RData")
load("R/nov02_results.RData")

set.seed(33359189)
source("R/figs4paper.R")
```

We can again use the trick of reading in the code, commenting out the
`postscript()` and `dev.off()`, and running it, to have the figures
appear within the present document. Here is Figure 1a.

```{r figure1a_again, fig.width=6, fig.height=6}
rm(list=ls())
load("R/sept02_data.RData")
load("R/nov02_data.RData")
load("R/nov02_results.RData")

figs <- readLines("R/figs4paper.R")

# comment out postscript() and dev.off()
figs <- sub("postscript(", "#postscript(", figs, fixed=TRUE)
figs <- sub("dev.off(", "#dev.off(", figs, fixed=TRUE)

# run the code here, just for figure 1a
eval(parse(text=figs[1:13]))
```

This looks the same as the [published figure](https://bit.ly/lamichhane2003_fig1).

And here is Figure 2.

```{r figure2_again, fig.width=10, fig.height=7}
set.seed(33359189)
eval(parse(text=figs[-(1:13)]))
```

Again, it looks the same as the [published figure](https://bit.ly/lamichhane2003_fig2).

### Verify Figure 2 results

We can verify the detailed results in Figure 2 by comparing the
present results to those that were saved in the
[original project
repository](https://github.com/kbroman/Project_Lamichhane2003),
included in the present repository as
[`Nov02/R/finalresults.rds`](https://github.com/kbroman/Paper_ReScience2020/blob/master/original/Nov02/R/finalresults.rds).

Here is a plot of the differences (original - reconstructed) in the
estimated probabilities of genes being essential.

```{r verify_fig2_results}
load("R/nov02_results.RData")
finalres_orig <- readRDS("../original/Nov02/R/finalresults.rds")

par(las=1, mar=c(4.1, 4.1, 0.6, 0.6))
grayplot(numTAs,
         (finalres_orig$bygene - finalres$bygene)*100,
         xlab="Number of TAs in proximal portion of gene",
         ylab="Difference in % probability (original - reproduction)")
```

The median difference is `r (med <- round(median( (finalres_orig$bygene - finalres$bygene)*100 ), 2))`.
So the reconstructed probabilities in Figure 2 have changed < 0.1%, which is immaterial and
likely due to MCMC sampling error.

Let me clarify: the positive values in Figure 2 are like
0.35&ndash;0.95, and we have reproduced them to within about 0.001. As
percentages, they are like 35-95%, and we have got them nailed down to
about 0.1%.

I'll also save this figure as a PDF for the paper.

```{r plot_fig2_diffs}
pdf("Figs/fig2_diff.pdf", height=4.5, width=6.5, pointsize=10)
<<verify_fig2_results>>
dev.off()
```

## Summary

The main difficulties in reproducing the results in the paper concerned poor
documentation and organization of the project data and code.

- I should have included a ReadMe file that described the structure of
  the project.

- I should have revised the scripts to give just the final published
  analyses, rather than create multiple mutated copies with some
  results pulled from one set of scripts and other results pulled from other
  scripts.

- I should have saved the random number generator seeds.

- I should have saved the key intermediate results to files and then
  loaded the files at the top of
  scripts that needed them, rather than rely on particular objects
  being present in the R environment.

- I should have indicated (and loaded) the required R libraries.

Nevertheless, I was able to reconstruct the results to within a small
and immaterial level of MCMC sampling error. I needed to make just
a few small changes to the code.

- Cutoffs for determining which genes and gene families to display in
  Tables 3 and 4 needed to be modified slightly.

- One vector that was derived using the R function `table()` needed to
  be converted from class `"table"` to a simple numeric vector. This
  is due to some change in R between version 1.5.1 (July 2002) and
  3.6.2 (current).

I was *not* able to reproduce Figure 3, however. The simulation
results and the code to create the figure are available, but I could
not find the code that was used to perform the simulations.

## Session info

Here is information about the version of R and R packages that I used.

```{r session_info}
devtools::session_info()
```

### 2020-07-14 response.md
Added response.md.

@karthik @kbroman Leah Wang (@lwa19) and I enjoyed going through
Karl’s [report][article] and [code][repo], and reading about his
experiences trying to reproduce the results of a paper written (well
over!) 10 years ago. As you may know, we (together with @jdblischak
and @stephens999) have been developing the
[workflowr R package][workflowr-pkg] (see also
[here][workflowr-paper]) to help researchers develop more reproducible
code. Karl’s remarks (see “Lessons”) suggest that workflowr could have
prevented — or at the very least could have alleviated — several of
the difficulties he encountered; in particular, workflowr automates or
facilitiates setting the random seed, maintaining a project history
(git), generating reports (R Markdown), and better project
organization. Our comments here focus on our experiencing using Karl’s
code to reproduce his results.

First, and foremost, we were able to successfully reproduce
[Karl’s results][reproduction]. The instructions are sufficiently
detailed and easy to follow. We ran the code on the high-performance
computing cluster run by the University of Chicago’s Research
Computing Center; see
[the fork of Karl’s repository for details][repo-fork], and in
particular [our copy of the report][reproduction-copy].

Please assign most of the credit for this review to Leah Wang
(@lwa19). She worked through Karl’s [report][reproduction], reproduced
his results and provided most of the feedback on the code.

[article]: https://kbroman.org/Paper_ReScience2020/article/article.pdf
[repo]: https://github.com/kbroman/Paper_ReScience2020
[repo-fork]: https://github.com/lwa19/Paper_ReScience2020
[reproduction]: https://kbroman.org/Paper_ReScience2020/reproduction/reproduction.html
[reproduction-copy]: https://lwa19.github.io/Paper_ReScience2020/reproduction/reproduction.html
[workflowr-pkg]: https://cran.r-project.org/package=workflowr
[workflowr-paper]: https://doi.org/10.12688/f1000research.20843.1

### 2020-07-14 response.md
Added more content to response.md.

@@
 [workflowr R package][workflowr-pkg] (see also
-[here][workflowr-paper]) to help researchers develop more reproducible
-code. Karl’s remarks (see “Lessons”) suggest that workflowr could have
-prevented — or at the very least could have alleviated — several of
-the difficulties he encountered; in particular, workflowr automates or
-facilitiates setting the random seed, maintaining a project history
-(git), generating reports (R Markdown), and better project
-organization. Our comments here focus on our experiencing using Karl’s
-code to reproduce his results.
+[the paper][workflowr-paper]) to help researchers develop more
+reproducible code. Karl’s remarks at the end of his article
+("Lessons") suggest that workflowr could have prevented --- or, at the
+very least, could have alleviated --- several of these difficulties;
+in particular, workflowr facilitiates, or automates, setting the
+random seed, maintaining a project history (with git), generating
+reports (with R Markdown), and better project organization. Our
+comments here focus on our experiences using Karl’s code to reproduce
+his results.
 
@@
 [Karl’s results][reproduction]. The instructions are sufficiently
-detailed and easy to follow. We ran the code on the high-performance
+detailed and easy to follow; in particular, it was helpful to have a
+single R Markdown file giving a start-to-finish set of steps for
+generating the results. We ran the code on the high-performance
 computing cluster run by the University of Chicago’s Research
-Computing Center; see
-[the fork of Karl’s repository for details][repo-fork], and in
-particular [our copy of the report][reproduction-copy].
+Computing Center; see [Leah's fork of Karl’s repository][repo-fork]
+for details, and in particular
+[Leah's reproduction of the report][reproduction-copy]. The only issue
+was that there was some knitr-specific syntax in the code chunks,
+e.g., `<<run_sept02_mcmc>>`, that gave RStudio some trouble depending
+on how the code was run. For compatibility it would be best to avoid
+such syntax. The remaining comments are some areas where we found the
+code difficult to follow, with some small suggestions for improvement
+(say, for the 50-year Reproducibility Challenge):
+
++ Changing the working directory multiple times during the analysis
+  makes it hard to keep track of the file paths, which files are being
+  loaded, and where files are being saved to, certainly when the code
+  is being run interactively; it would be better to have a single
+  working directory throughout, and define all the file paths relative
+  to that directory.
+
++ In the "table4" code chunk was rather headspinning in the way that
+  the code was modified on the fly; it seems that this could have been
+  handdled more elegantly by defining a variable, e.g., `cutoff <-
+  75`, and using that instead of hardcoded values.
+
++ Instead of clearing the workspace with `rm(list = ls())`, it may
+  have been better to split the analysis into multiple R Markdown
+  files that are listed in a "master" R Markdown file.
+
++ Since the MCMC step takes a some time to run, it would be helpful to
+  give a warning, and some sense of how long it should take to run.
+
++ There is some bookkeeping at the beginning---moving and renaming files---and perhaps it would have been better to commit these changes..
+
++ All the code and results are saved to a single folder,
+  `Paper_ReScience2020/reproduction/R`. This is what we see:
+
+    ```
+    circlefig.R
+    figs4paper.R
+    nov02_analysis.R
+    nov02_data.RData
+    nov02_prepareData.R
+    nov02_results.RData
+    prob_essential.ps
+    sept02_analysis.R
+    sept02_data.RData
+    sept02_prepareData.R
+    sept02_results.RData
+    ```
+
+  It would have been helpful to create folders such as "code", "data"
+  and "output" to distinguish the inputs from the outputs.
 

### 2020-07-14 response.md
Added more stuff to response.md.

@@
 Computing Center; see [Leah's fork of Karl’s repository][repo-fork]
-for details, and in particular
-[Leah's reproduction of the report][reproduction-copy]. The only issue
-was that there was some knitr-specific syntax in the code chunks,
-e.g., `<<run_sept02_mcmc>>`, that gave RStudio some trouble depending
-on how the code was run. For compatibility it would be best to avoid
-such syntax. The remaining comments are some areas where we found the
-code difficult to follow, with some small suggestions for improvement
-(say, for the 50-year Reproducibility Challenge):
+for details, and in particular [Leah's reproduction of the
+report][reproduction-copy]. The only issue was that there was some
+knitr-specific syntax in the code chunks, e.g., `<<run_sept02_mcmc>>`,
+that gave RStudio some trouble depending on how the code was run. For
+compatibility it would be best to avoid such syntax. The remaining
+comments are some areas where we found the code difficult to follow,
+with some small suggestions for improvement (say, for the 50-year
+Reproducibility Challenge):
 
@@
 
-+ There is some bookkeeping at the beginning---moving and renaming files---and perhaps it would have been better to commit these changes..
++ There is some bookkeeping at the beginning---moving and renaming
+  files---and perhaps it would have been better to commit these
+  changes to the git repository, and not include them in the script.
 
@@
 
-Please assign most of the credit for this review to Leah Wang
-(@lwa19). She worked through Karl’s [report][reproduction], reproduced
-his results and provided most of the feedback on the code.
++ When you said, "Here is the reconstruction of Table 3", what you
+  really meant, we believe, is Supplement Table 6," which is the same
+  as Table 3, but with the longer list of genes. There were a couple
+  spots where the explanations weren't very clear, at least without
+  having gone through the remainder of the analysis; please see
+  [the comments included in the R Markdown][reproduction-copy].
+
+Please assign most of the credit for this review to Leah Wang (@lwa19).
+She worked through Karl’s [report][reproduction], reproduced his
+results and provided most of the feedback on the code.
 

### 2020-07-14 response.md
Some edits to response.md.

@@
 [workflowr R package][workflowr-pkg] (see also
-[the paper][workflowr-paper]) to help researchers develop more
-reproducible code. Karl’s remarks at the end of his article
-("Lessons") suggest that workflowr could have prevented --- or, at the
-very least, could have alleviated --- several of these difficulties;
+[the workflowr paper][workflowr-paper]) to help researchers develop
+more reproducible code, and Karl’s remarks at the end of his article
+("Lessons") suggest that workflowr could have prevented—or, at the
+very least, could have alleviated—several of the difficulties;
 in particular, workflowr facilitiates, or automates, setting the
@@
 that gave RStudio some trouble depending on how the code was run. For
-compatibility it would be best to avoid such syntax. The remaining
-comments are some areas where we found the code difficult to follow,
-with some small suggestions for improvement (say, for the 50-year
-Reproducibility Challenge):
+cross-compatibility perhaps this syntax is best avoided. The remaining
+comments touch on are some areas where we found the code difficult to
+follow, and include some small suggestions for improvement (say, with
+a 50-year Reproducibility Challenge in mind):
 
@@
   loaded, and where files are being saved to, certainly when the code
-  is being run interactively; it would be better to have a single
-  working directory throughout, and define all the file paths relative
+  is being run interactively. It would be better to have a single
+  working directory throughout and define all the file paths relative
   to that directory.
 
-+ In the "table4" code chunk was rather headspinning in the way that
-  the code was modified on the fly; it seems that this could have been
-  handdled more elegantly by defining a variable, e.g., `cutoff <-
-  75`, and using that instead of hardcoded values.
++ The "table4" code chunk was particularly confusing because of the
+  way that the code was modified on the fly; it seems that this could
+  have been handled more simply by defining a variable, e.g., `cutoff
+  <- 75`, and using that instead of hardcoded values.
 
@@
 
-+ Since the MCMC step takes a some time to run, it would be helpful to
-  give a warning, and some sense of how long it should take to run.
++ Since the MCMC step took a some time to run, it would have been
+  helpful to give a warning, and some sense of expected runtime.
 
@@
   It would have been helpful to create folders such as "code", "data"
-  and "output" to distinguish the inputs from the outputs.
+  and "output" to make clear which are the inputs and which are the
+  outputs.
 
@@
   as Table 3, but with the longer list of genes. There were a couple
-  spots where the explanations weren't very clear, at least without
-  having gone through the remainder of the analysis; please see
+  spots where the explanations weren't clear, at least without
+  being familiar with later steps in analysis; please also see
   [the comments included in the R Markdown][reproduction-copy].
